Closure-Free Functional Programming in a Two-Level Type Theory

### 2024-07-18 icfp24paper/paper.tex
camera ready
stuff

\title{Closure-Free Functional Programming in a Two-Level Type Theory}
\titlenote{This work was supported
by grant 2019.0116 of the Knut and Alice Wallenberg Foundation.}

@@

\subsubsection{Comparison to defunctionalization}
Defunctionalization is a program translation which represents higher-order
functions using only first-order functions and inductive data
\cite{DBLP:journals/lisp/Reynolds98a,DBLP:conf/ppdp/DanvyN01}. The idea is to
represent each closure as a constructor of an inductive type, and implement
closure application as a top-level first-order function which switches on all
possible closure constructors. Hence, like our unstaging algorithm,
defunctionalization produces first-order code from higher-order code. We summarize
the differences.
\begin{itemize}
\item Defunctionalization mainly supports \emph{compiler optimizations}, by making
  closures transparent to analysis. It is not an optimization by itself; it does
  not change the amount of dynamic control flow or dynamic allocations in a
  program. A dynamic closure call becomes a dynamic case-switch and a
  heap-allocated ``functional'' closure becomes a heap-allocated inductive
  constructor.
\item Staging mainly supports optimization by \emph{programmers} by providing
  control over code generation. Unstaging does not translate higher-order
  functions to anything in particular. Instead, unstaging is the execution of
  higher-order metaprograms which produce first-order programs.
\end{itemize}

@@

\subsubsection{Case splitting in monads}
We often want to case-split on object-level data inside a monadic action, and
perform different actions in different branches. At first, this may seem
problematic, because we cannot directly compute metaprograms from object-level
case splits. Fortunately, with a little bit more work, this is possible in any
$\MonadGen$, for any value type, using a variation of the so-called ``trick''
\cite{eta-expansion-trick}.

@@

\subsubsection{$\State$ and $\Reader$ operations} If we use
the $\modify$ function that we already have in $\State_\M$, a curious thing
happens. The meaning of $\modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx})$ is to
replace the current state $\vx$, as an object expression, with the expression
$\qt{\spl \vx + \spl \vx}$, and this happens at unstaging time. This behaves as an
``inline'' modification which replaces every subsequent mention of the state
with a different expression. For instance, ignoring newtype wrappers for now,
\[ \mdown \dlr \mdo \{\modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx}); \modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx});\return\,\qt{()}\} \]
is unstaged to
\[ \qt{\lam \vx.\,((),\,(\vx + \vx) + (\vx + \vx))} \]
which duplicates the evaluation of $\vx + \vx$. The duplication can be avoided
by let-binding the result in the object language. A similar phenomenon happens
with the $\mit{local}$ function in $\Reader$. So we define ``stricter'' versions
of these operations. We also return $\Up ()$ from actions instead of $()$ ---
the former is more convenient, because the $\mdown$ operation can be immediately
used on it.
\begingroup
\allowdisplaybreaks
\begin{alignat*}{3}
  & \mput' : (\MonadState\,(\Up \vS)\,\vM,\,\MonadGen\,\vM) \RA \Up \vS \to \vM\,(\Up ()) \\
  & \mput'\,\vs = \mdo \{\vs \fro \gen\,\vs; \mput\,\vs; \return \qt{()}\}\\
  &\\
  & \modify' : (\MonadState\,(\Up \vS)\,\vM,\,\MonadGen\,\vM) \RA (\Up \vS \to \Up \vS) \to \vM\,(\Up ()) \\
  & \modify'\,\vf = \mdo \{\vs \fro \get; \mput'\,(\vf\,\vs)\}\\
  &\\
  & \mit{local'} : (\MonadReader\,(\Up \vR)\,\vM,\,\MonadGen\,\vM) \RA (\Up \vR \to \Up \vR) \to \vM\,\vA \to \vM\,\vA\\
  & \mit{local'}\,\vf\,\vma = \mdo \{\vr \fro \mit{ask}; \vr \fro \gen\,(\vf\,\vr);\mit{local}\,(\lam \_.\,\vr)\,\vma\}
\end{alignat*}
\endgroup
Now,
\[ \mdown \dlr \mdo \{\modify'\,(\lam \vx.\,\qt{\spl \vx + \spl \vx}); \modify'\,(\lam \vx.\,\qt{\spl \vx + \spl \vx})\} \]
is unstaged to
\[ \qt{\lam \vx.\,\letdef x := \vx + \vx; \letdef \vx := \vx + \vx; ((),\,\vx)}. \]

@@

There is a deficiency in our library so far. Consider:
\begin{alignat*}{3}
  & \vf : \Bool \to \StateT\,\Nat\,(\MaybeT\,\Identity)\,()\\
  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
  & \ind \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20}\\
  & \ind\modify'\,(\lambda\,x.\,\qt{\spl x + \spl x}))
\end{alignat*}
This is unstaged to the following (omitting newtype wrappers):
\begin{alignat*}{3}
  & \vf\,\vb\,\vs := \case\,\vb\,\of \\
  & \ind \True\hspace{0.2em} \to \letdef \vs := 10; \letdef \vs := \vs + \vs; \Just\,((),\vs)\\
  & \ind \False \to \letdef \vs := 20; \letdef \vs := \vs + \vs; \Just\,((),\vs)
\end{alignat*}
Notice that the final state modification gets inlined into both branches. This
duplication follows from the definition of monadic binding in $\Gen$ and the
$\mit{split}$ function in the desugaring of $\mbf{case}$. Code generation is
continued in both branches with the same action. If we have multiple Boolean
$\mbf{case}$ splits sequenced after each other, that yields exponential code
size. A possible fix is to let-bind the branching action:
\begin{alignat*}{3}
  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
  & \ind \msf{act} \fro \gen \dlr \mdown \dlr \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20} \\
  & \ind \mup\,\msf{act} \\
  & \ind\modify\,(\lambda\,x.\,\qt{\spl x + \spl x}))
\end{alignat*}
However, this yields suboptimal code:
\begin{alignat*}{3}
  & \vf\,\vb\,\vs :=\\
  & \ind \letdef\,\msf{act}\,\vs := \case\,\vb\,\of \True\hspace{0.2em}\to \letdef \vs := 10; \Just\,((), \vs)\\
  & \ind\hspace{8.5em}                               \False \to \letdef \vs := 20; \Just\,((), \vs);\\
  & \ind \case \msf{act}\,\vs\,\of\\
  & \ind\ind \Just\,(\_,\,\vs) \to \letdef \vs := \vs + \vs; \Just\,((),\,\vs)\\
  & \ind\ind \Nothing\hspace{0.6em} \to \Nothing
\end{alignat*}
This solution is fairly good when we only have $\State$ or $\Reader$ effects, where
$\mdown$ only introduces a runtime pair constructor, and it is feasible to
compile object-level pairs as unboxed data, without overheads. However, for
$\Maybe$, $\mdown$ introduces a runtime $\Just$ or $\Nothing$, and $\mup$
introduces a runtime case split. A better solution would be to introduce two
let-bound join points before the offending $\mbf{case}$, one for returning a
$\Just$ and one for returning $\Nothing$, but fusing away the actual runtime
constructors:
\begin{alignat*}{3}
  & \vf\,\vb\,\vs :=\\
  & \ind \letdef\,\msf{joinNothing}\,\vs := \Nothing;\\
  & \ind \letdef\,\msf{joinJust}\,\hspace{1.8em}\vs := \letdef \vs := \vs + \vs; \Just\,((),\,\vs);\\
  & \ind \case \msf{b}\,\of\\
  & \ind \ind \True\hspace{0.2em}  \to \letdef \vs := 10; \msf{joinJust}\,\vs\\
  & \ind \ind \False \to \letdef \vs := 20; \msf{joinJust}\,\vs
\end{alignat*}
Here, $\msf{joinNothing}$ happens to be dead code, but it is easy to clean up in
downstream compilation. Such ``fused'' returns are possible whenever we have a
$\Gen\,\vA$ action at the bottom of the transformer stack, such that $\vA$ is
isomorphic to a meta-level finite sum of value types. Recall that $\Gen\,\vA$ is
defined as $\{\vR : \VTy\} \to (\vA \to \Up \vR) \to \Up \vR$. Here, if $\vA$ is
a finite sum, we can rearrange $\vA \to \Up \vR$ to a finite product of
functions.

@@

Accordingly, we define an isomorphic presentation of $\El_\SOP\,\vA \to
\Lift \vR$ as a product of object-level functions:
\begin{alignat*}{3}
  &\mathrlap{\mit{Fun}_{\SOP\Lift} : \USOP \to \Ty \to \MTy}\\
  &\mit{Fun}_{\SOP\Lift}\,\Nil\,&& \vR = ()\\
  &\mit{Fun}_{\SOP\Lift}\,(\Cons\,\vA\,\vB)\,&& \vR = (\Lift(\mit{foldr}\,(\to)\,\vR\,\vA),\,\mit{Fun}_{\SOP\Lift}\,\vB\,\vR)
\end{alignat*}
\begin{alignat*}{3}
  &\mtabulate &&: (\El_\SOP\,\vA \to \Lift \vR) \to \mit{Fun}_{\SOP\Lift}\,\vA\,\vR\\
  &\mindex    &&: \mit{Fun}_{\SOP\Lift}\,\vA\,\vR \to (\El_\SOP\,\vA \to \Lift \vR)
\end{alignat*}
We omit here the definitions of $\mtabulate$ and $\mindex$. We will also need to
let-bind all functions in a $\mit{Fun}_{\SOP\Lift}$:
\begin{alignat*}{3}
  & \mathrlap{\mit{genFun}_{\SOP\Lift} : \{\vA : \USOP\} \to \mit{Fun}_{\SOP\Lift}\,\vA\,\vR \to \mit{Fun}_{\SOP\Lift}\,\vA\,\vR}\\
  & \mit{genFun}_{\SOP\Lift}\,\{\Nil\}        && ()             &&= \return\,()\\
  & \mit{genFun}_{\SOP\Lift}\,\{\Cons\,\_\,\vA\}&& (\vf,\,\mit{\vfs}) &&= ({,})\,{\fmap}\,\gen\,\vf\,{\ap}\,\mit{genFun}_{\SOP\Lift}\,\{\vA\}\,\vfs
\end{alignat*}
We introduce a class for monads that support control flow joining.
\begin{alignat*}{3}
  & \tyclass\,\Monad\,\vM \RA \MonadJoin\,\vM\,\where\\
  & \ind \join : \IsSOP\,\vA \RA \vM\,\vA \to \vM\,\vA
\end{alignat*}
The most interesting instance is the ``base case'' for $\Gen$:
\begin{alignat*}{3}
  &\instance \MonadJoin\,\Gen\,\where\\
  &\ind \join\,\vma = \Gen \dlr \lam \vk.\,\runGen \dlr \mdo\\
  &\ind \ind \mit{joinPoints} \fro \mit{genFun}_{\SOP\Lift}\,(\mtabulate\,(\vk \circ \mit{rep}^{-1}))\\
  &\ind \ind \va \fro \vma\\
  &\ind \ind \return \dlr \mindex\,\mit{joinPoints}\,(\mit{rep}\,\va)
\end{alignat*}
Here we first convert $\vk : \vA \to \Lift \vR$ to a product of join points and
let-bind each one of them. Then we generate code that returns to the appropriate
join point for each return value. Other $\MonadJoin$ instances are straightforward. In
$\StateT_\M$, we need the extra $\IsSOP\,\vS$ constraint because $\vS$ is returned
as a result value.
\begin{alignat*}{3}
  & \instance\,(\MonadJoin\,\vM) \RA \MonadJoin\,(\MaybeT_\M\,\vM)\,\where\\
  & \ind \join\,(\MaybeT_\M\,\vma) = \MaybeT_\M\,(\join\,\vma)\\
  & \instance\,(\MonadJoin\,\vM) \RA \MonadJoin\,(\ReaderT_\M\,\vR\,\vM)\,\where\\
  & \ind \join\,(\ReaderT_\M\,\vma) = \ReaderT_\M\,(\join \circ \vma)\\
  & \instance\,(\MonadJoin\,\vM,\,\IsSOP\,\vS) \RA \MonadJoin\,(\StateT_\M\,\vS\,\vM)\,\where\\
  & \ind \join\,(\StateT_\M\,\vma) = \StateT_\M\,(\join \circ \vma)
\end{alignat*}
Now, the following definition yields the previously seen code output with
$\msf{joinNothing}$ and $\msf{joinJust}$.
\begin{alignat*}{3}
  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
  & \ind \join \dlr \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20}\\
  & \ind\modify'\,(\lambda\,x.\,\qt{\spl x + \spl x}))
\end{alignat*}
It might make sense to also have a simple ``desugaring'' rule that inserts a
$\join$ whenever we have an object-level case split with more than one branch. At
worst, this generates some noise and dead code that is easy to remove by
conservative code optimization.

@@

\subsection{Running Streams with Mutual Recursion}

@@

While $\Pull$ is not a monad, and hence also not a $\MonadGen$, we can still use
a monadic style of stream programming with good ergonomics. First, we need
singleton streams for ``returning'':
\begin{alignat*}{3}
  & \mit{single} : \vA \to \Pull\,\vA \\
  & \mit{single}\,\va = \Pull\,\Bool_\M\,\True \dlr \lam \vb.\,\return \dlr\mbf{if}\,\vb\,\mbf{then}\,\Yield\,\va\,\False\,\mbf{else}\,\Stop
\end{alignat*}
Now, we should use this operation with care, since it has two states and can
contribute to a code size blow-up. For example, $\concatMap$ over $\single$
introduces at least a doubling of the number of state shapes. Although
let-insertion and case splitting could be derived as $\concatMap$ over
$\single$, to avoid the size explosion we give more specialized definitions
instead. First, we define an embedding of $\Gen$ operations in streams.
\begin{alignat*}{3}
  & \hspace{-11em}\mathrlap{\mit{bind_{\Gen}} : \IsSOP\,\vA \RA \Gen\,\vA \to (\vA \to \Pull\,\vB) \to \Pull\,\vB}\\
  & \hspace{-11em}\mathrlap{\mit{bind_{\Gen}}\,\vga\,\vf\,=}\\
  & \hspace{-11em}\mathrlap{\ind \Pull\,(\Sigma\,\vA\,(\mit{projS} \circ \vf))\,(\mdo \{\va \fro \vga; \vs \fro \mit{projSeed}\,(\vf\,\va); \return (\va,\,\vs)) \dlr \lam \case}\\
  & \hspace{-11em}\mathrlap{\ind \ind (\va,\,\vs) \to \mit{projStep}\,(\vf\,\va)\,\vs \bind \lam \case }\\
  & \hspace{-11em}\ind \ind \ind \Stop            &&\to \return \Stop\\
  & \hspace{-11em}\ind \ind \ind \Skip\,\vs       &&\to \return \dlr \Skip\,(\va,\,\vs) \\
  & \hspace{-11em}\ind \ind \ind \Yield\,\vb\,\vs &&\to \return \dlr \Yield\,\vb\,(\va,\,\vs)
\end{alignat*}
We recover let-insertion and case splitting as follows:
\begin{alignat*}{3}
  &\mit{gen_\Pull} : \Lift \vA \to (\Lift \vA \to \Pull\,\vB) \to \Pull\,\vB\\
  &\mit{gen_\Pull}\,\va = \mit{bind_{\Gen}}\,(\gen\,\va)
\end{alignat*}
\begin{alignat*}{3}
  &\mit{case_\Pull} : (\mit{Split}\,\vA,\,\IsSOP\,(\mit{SplitTo}\,\{\vA\})) \RA \Lift \vA \to (\mit{SplitTo}\,\{\vA\} \to \Pull\,\vB) \to \Pull\,\vB\\
  &\mit{case_\Pull}\,\va = \mit{bind_{\Gen}}\,(\mit{split}\,\va)
\end{alignat*}
Let us look at small example. We write $\mit{forEach}$ for $\concatMap$ with
flipped arguments.
\begin{alignat*}{3}
  &\hspace{-4em}\mathrlap{\mit{forEach}\,(\mit{take}\,\qt{100}\,(\mit{countFrom}\,\qt{0})) \dlr \lam \vx.}\\
  &\hspace{-4em}\mathrlap{\gen_\Pull\, \qt{\spl \vx * 2} \dlr \lam \vy.}\\
  &\hspace{-4em}\mathrlap{\mit{case_\Pull}\,\qt{\spl \vx < 50} \dlr \lam \case}\\
  &\hspace{-4em} \ind \True  &&\to \mit{take}\,\vy\,(\mit{countFrom}\,x)\\
  &\hspace{-4em} \ind \False &&\to \single\,\vy
\end{alignat*}
Here, in every $\forEach$ iteration, $\gen_\Pull\,\qt{\spl \vx * 2}$ evaluates
the given expression and saves the result to the machine state. Then,
$\mit{case_\Pull}$ branches on an object-level Boolean expression. If we use
$\foldr$ to generate object code from this definition, we get \emph{four}
mutually defined functions. We get two state shapes from $\single\,\vy$ and one
from $\mit{take}\,\vy\,(\mit{countFrom}\,x)$; we sum these to get three for the
$\mit{case_\Pull}$, then finally we get an extra shape from $\mit{forEach}$ which
introduces an additional $\Maybe$.

@@

\section*{Acknowledgements}
I thank the anonymous reviewers, Sebastian Graf and Jacques Carette, for their
comments and suggestions on draft versions of this paper.

@@

\subsection{Identity type}

>>>>

\title{Closure-Free Functional Programming in a Two-Level Type Theory}

@@

\subsubsection*{Comparison to defunctionalization}
Defunctionalization is a program translation which represents higher-order
functions using only first-order functions and inductive data
\cite{DBLP:journals/lisp/Reynolds98a,DBLP:conf/ppdp/DanvyN01}. The idea is to
represent each closure as a constructor of an inductive type, and implement
closure application as a top-level first-order function which switches on all
possible closure constructors. Hence, like our unstaging algorithm,
defunctionalization produces first-order code from higher-order code. We summarize
the differences.
\begin{itemize}
\item Defunctionalization mainly supports \emph{compiler optimizations}, by making
  closures transparent to analysis. It is not an optimization by itself; it does
  not change the amount of dynamic control flow or dynamic allocations in a
  program. A dynamic closure call becomes a dynamic case-switch and a
  heap-allocated ``functional'' closure becomes a heap-allocated inductive
  constructor.
\item Staging mainly supports optimization by \emph{programmers} by providing
  control over code generation. Unstaging does not translate higher-order
  functions to anything in particular. Instead, unstaging is the execution of
  higher-order metaprograms which produce first-order programs.
\end{itemize}

@@

\subsubsection*{Case splitting in monads}
We often want to case-split on object-level data inside a monadic action, and
perform different actions in different branches. At first, this may seem
problematic, because we cannot directly compute metaprograms from object-level
case splits. Fortunately, with a little bit more work, this is possible in any
$\MonadGen$, for any value type, using a variation of the so-called ``trick''
\cite{eta-expansion-trick}.

@@

\subsubsection*{$\State$ and $\Reader$ operations} If we use
the $\modify$ function that we already have in $\State_\M$, a curious thing
happens. The meaning of $\modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx})$ is to
replace the current state $\vx$, as an object expression, with the expression
$\qt{\spl \vx + \spl \vx}$, and this happens at unstaging time. This behaves as an
``inline'' modification which replaces every subsequent mention of the state
with a different expression. For instance, ignoring newtype wrappers for now,
\[ \mdown \dlr \mdo \{\modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx}); \modify\,(\lam \vx.\,\qt{\spl \vx + \spl \vx});\return\,\qt{()}\} \]
is unstaged to
\[ \qt{\lam \vx.\,((),\,(\vx + \vx) + (\vx + \vx))} \]
which duplicates the evaluation of $\vx + \vx$. The duplication can be avoided
by let-binding the result in the object language. A similar phenomenon happens
with the $\mit{local}$ function in $\Reader$. So we define ``stricter'' versions
of these operations. We also return $\Up ()$ from actions instead of $()$ ---
the former is more convenient, because the $\mdown$ operation can be immediately
used on it.
\begingroup
\allowdisplaybreaks
\begin{alignat*}{3}
  & \mput' : (\MonadState\,(\Up \vS)\,\vM,\,\MonadGen\,\vM) \RA \Up \vS \to \vM\,(\Up ()) \\
  & \mput'\,\vs = \mdo \{\vs \fro \gen\,\vs; \mput\,\vs; \return \qt{()}\}\\
  &\\
  & \modify' : (\MonadState\,(\Up \vS)\,\vM,\,\MonadGen\,\vM) \RA (\Up \vS \to \Up \vS) \to \vM\,(\Up ()) \\
  & \modify'\,\vf = \mdo \{\vs \fro \get; \mput'\,(\vf\,\vs)\}\\
  &\\
  & \mit{local'} : (\MonadReader\,(\Up \vR)\,\vM,\,\MonadGen\,\vM) \RA (\Up \vR \to \Up \vR) \to \vM\,\vA \to \vM\,\vA\\
  & \mit{local'}\,\vf\,\vma = \mdo \{\vr \fro \mit{ask}; \vr \fro \gen\,(\vf\,\vr);\mit{local}\,(\lam \_.\,\vr)\,\vma\}
\end{alignat*}
\endgroup
Now,
\[ \mdown \dlr \mdo \{\modify'\,(\lam \vx.\,\qt{\spl \vx + \spl \vx}); \modify'\,(\lam \vx.\,\qt{\spl \vx + \spl \vx})\} \]
is unstaged to
\[ \qt{\lam \vx.\,\letdef \vx := \vx + \vx; \letdef \vx := \vx + \vx; ((),\,\vx)}. \]

@@

There is a deficiency in our library so far. Consider:
\begin{alignat*}{3}
  & \vf : \Bool \to \StateT\,\Nat\,(\MaybeT\,\Identity)\,()\\
  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
  & \ind \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20}\\
  & \ind\modify'\,(\lambda\,\vx.\,\qt{\spl \vx + \spl \vx}))
\end{alignat*}
This is unstaged to the following (omitting newtype wrappers):
\begin{alignat*}{3}
  & \vf\,\vb\,\vs := \case\,\vb\,\of \\
  & \ind \True\hspace{0.2em} \to \letdef \vs := 10; \letdef \vs := \vs + \vs; \Just\,((),\vs)\\
  & \ind \False \to \letdef \vs := 20; \letdef \vs := \vs + \vs; \Just\,((),\vs)
\end{alignat*}
Notice that the final state modification gets inlined into both branches. This
duplication follows from the definition of monadic binding in $\Gen$ and the
$\mit{split}$ function in the desugaring of $\mbf{case}$. Code generation is
continued in both branches with the same action. If we have multiple Boolean
$\mbf{case}$ splits sequenced after each other, that yields exponential code
size. A possible fix is to let-bind the branching action:
\begin{alignat*}{3}
  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
  & \ind \msf{act} \fro \gen \dlr \mdown \dlr \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20} \\
  & \ind \mup\,\msf{act} \\
  & \ind\modify\,(\lambda\,\vx.\,\qt{\spl \vx + \spl \vx}))
\end{alignat*}
However, this yields suboptimal code:
\begin{alignat*}{3}
  & \vf\,\vb\,\vs :=\\
  & \ind \letdef\,\msf{act}\,\vs := \case\,\vb\,\of \True\hspace{0.2em}\to \letdef \vs := 10; \Just\,((), \vs)\\
  & \ind\hspace{8.5em}                               \False \to \letdef \vs := 20; \Just\,((), \vs);\\
  & \ind \case \msf{act}\,\vs\,\of\\
  & \ind\ind \Just\,(\_,\,\vs) \to \letdef \vs := \vs + \vs; \Just\,((),\,\vs)\\
  & \ind\ind \Nothing\hspace{0.6em} \to \Nothing
\end{alignat*}
This solution is fairly good when we only have $\State$ or $\Reader$ effects, where
$\mdown$ only introduces a runtime pair constructor, and it is feasible to
compile object-level pairs as unboxed data, without overheads. However, for
$\Maybe$, $\mdown$ introduces a runtime $\Just$ or $\Nothing$, and $\mup$
introduces a runtime case split. A better solution would be to introduce two
let-bound join points before the offending $\mbf{case}$, one for returning a
$\Just$ and one for returning $\Nothing$, but fusing away the actual runtime
constructors:
\begin{alignat*}{3}
  & \vf\,\vb\,\vs :=\\
  & \ind \letdef\,\msf{joinNothing}\,\vs := \Nothing;\\
  & \ind \letdef\,\msf{joinJust}\,\hspace{1.8em}\vs := \letdef \vs := \vs + \vs; \Just\,((),\,\vs);\\
  & \ind \case \msf{b}\,\of\\
  & \ind \ind \True\hspace{0.2em}  \to \letdef \vs := 10; \msf{joinJust}\,\vs\\
  & \ind \ind \False \to \letdef \vs := 20; \msf{joinJust}\,\vs
\end{alignat*}
Here, $\msf{joinNothing}$ happens to be dead code, but it is easy to clean up in
downstream compilation. Such ``fused'' returns are possible whenever we have a
$\Gen\,\vA$ action at the bottom of the transformer stack, such that $\vA$ is
isomorphic to a meta-level finite sum of value types. Recall that $\Gen\,\vA$ is
defined as $\{\vR : \VTy\} \to (\vA \to \Up \vR) \to \Up \vR$. Here, if $\vA$ is
a finite sum, we can rearrange $\vA \to \Up \vR$ to a finite product of
functions.

@@

Accordingly, we define an isomorphic presentation of $\El_\SOP\,\vA \to
\Lift \vR$ as a product of object-level functions:
\begin{alignat*}{3}
  &\mathrlap{\mit{Fun}_{\SOP\Lift} : \USOP \to \Ty \to \MTy}\\
  &\mit{Fun}_{\SOP\Lift}\,\Nil\,&& \vR = ()\\
  &\mit{Fun}_{\SOP\Lift}\,(\Cons\,\vA\,\vB)\,&& \vR = (\Lift(\mit{foldr}\,(\to)\,\vR\,\vA),\,\mit{Fun}_{\SOP\Lift}\,\vB\,\vR)
\end{alignat*}
\begin{alignat*}{3}
  &\mtabulate &&: (\El_\SOP\,\vA \to \Lift \vR) \to \mit{Fun}_{\SOP\Lift}\,\vA\,\vR\\
  &\mindex    &&: \mit{Fun}_{\SOP\Lift}\,\vA\,\vR \to (\El_\SOP\,\vA \to \Lift \vR)
\end{alignat*}
We omit here the definitions of $\mtabulate$ and $\mindex$. We will also need to
let-bind all functions in a $\mit{Fun}_{\SOP\Lift}$:
\begin{alignat*}{3}
  & \mathrlap{\mit{genFun}_{\SOP\Lift} : \{\vA : \USOP\} \to \mit{Fun}_{\SOP\Lift}\,\vA\,\vR \to \mit{Fun}_{\SOP\Lift}\,\vA\,\vR}\\
  & \mit{genFun}_{\SOP\Lift}\,\{\Nil\}        && ()             &&= \return\,()\\
  & \mit{genFun}_{\SOP\Lift}\,\{\Cons\,\_\,\vA\}&& (\vf,\,\mit{\vfs}) &&= ({,})\,{\fmap}\,\gen\,\vf\,{\ap}\,\mit{genFun}_{\SOP\Lift}\,\{\vA\}\,\vfs
\end{alignat*}
We introduce a class for monads that support control flow joining.
\begin{alignat*}{3}
  & \tyclass\,\Monad\,\vM \RA \MonadJoin\,\vM\,\where\\
  & \ind \join : \IsSOP\,\vA \RA \vM\,\vA \to \vM\,\vA
\end{alignat*}
The most interesting instance is the ``base case'' for $\Gen$:
\begin{alignat*}{3}
  &\instance \MonadJoin\,\Gen\,\where\\
  &\ind \join\,\vma = \Gen \dlr \lam \vk.\,\runGen \dlr \mdo\\
  &\ind \ind \mit{joinPoints} \fro \mit{genFun}_{\SOP\Lift}\,(\mtabulate\,(\vk \circ \mit{rep}^{-1}))\\
  &\ind \ind \va \fro \vma\\
  &\ind \ind \return \dlr \mindex\,\mit{joinPoints}\,(\mit{rep}\,\va)
\end{alignat*}
Here we first convert $\vk : \vA \to \Lift \vR$ to a product of join points and
let-bind each one of them. Then we generate code that returns to the appropriate
join point for each return value. Other $\MonadJoin$ instances are straightforward. In
$\StateT_\M$, we need the extra $\IsSOP\,\vS$ constraint because $\vS$ is returned
as a result value.
\begin{alignat*}{3}
  & \instance\,(\MonadJoin\,\vM) \RA \MonadJoin\,(\MaybeT_\M\,\vM)\,\where\\
  & \ind \join\,(\MaybeT_\M\,\vma) = \MaybeT_\M\,(\join\,\vma)\\
  & \instance\,(\MonadJoin\,\vM) \RA \MonadJoin\,(\ReaderT_\M\,\vR\,\vM)\,\where\\
  & \ind \join\,(\ReaderT_\M\,\vma) = \ReaderT_\M\,(\join \circ \vma)\\
  & \instance\,(\MonadJoin\,\vM,\,\IsSOP\,\vS) \RA \MonadJoin\,(\StateT_\M\,\vS\,\vM)\,\where\\
  & \ind \join\,(\StateT_\M\,\vma) = \StateT_\M\,(\join \circ \vma)
\end{alignat*}
Now, the following definition yields the previously seen code output with
$\msf{joinNothing}$ and $\msf{joinJust}$.
\begin{alignat*}{3}
  & \vf\,\vb := \spl(\mdown \dlr \mdo \\
  & \ind \join \dlr \case \qt{\vb}\,\of\,\True\to \mput' \qt{10}; \False \to \mput' \qt{20}\\
  & \ind\modify'\,(\lambda\,\vx.\,\qt{\spl \vx + \spl \vx}))
\end{alignat*}
It might make sense to also have a simple ``desugaring'' rule that inserts a
$\join$ whenever we have an object-level case split with more than one branch. At
worst, this generates some noise and dead code that is easy to remove by
conservative code optimization.

@@

\subsection{Running Streams With Mutual Recursion}

@@

While $\Pull$ is not a monad, and hence also not a $\MonadGen$, we can still use
a monadic style of stream programming with good ergonomics. First, we need
singleton streams for ``returning'':
\begin{alignat*}{3}
  & \mit{single} : \vA \to \Pull\,\vA \\
  & \mit{single}\,\va = \Pull\,\Bool_\M\,\True \dlr \lam \vb.\,\return \dlr\mbf{if}\,\vb\,\mbf{then}\,\Yield\,\va\,\False\,\mbf{else}\,\Stop
\end{alignat*}
Now, we should use this operation with care, since it has two states and can
contribute to a code size blow-up. For example, $\concatMap$ over $\single$
introduces at least a doubling of the number of state shapes. Although
let-insertion and case splitting could be derived as $\concatMap$ over
$\single$, to avoid the size explosion we give more specialized definitions
instead. First, we define an embedding of $\Gen$ operations in streams.
\begin{alignat*}{3}
  & \hspace{-11em}\mathrlap{\mit{bind_{\Gen}} : \IsSOP\,\vA \RA \Gen\,\vA \to (\vA \to \Pull\,\vB) \to \Pull\,\vB}\\
  & \hspace{-11em}\mathrlap{\mit{bind_{\Gen}}\,\vga\,\vf\,=}\\
  & \hspace{-11em}\mathrlap{\ind \Pull\,(\Sigma\,\vA\,(\mit{projS} \circ \vf))\,(\mdo \{\va \fro \vga; \vs \fro \mit{projSeed}\,(\vf\,\va); \return (\va,\,\vs)\}) \dlr \lam \case}\\
  & \hspace{-11em}\mathrlap{\ind \ind (\va,\,\vs) \to \mit{projStep}\,(\vf\,\va)\,\vs \bind \lam \case }\\
  & \hspace{-11em}\ind \ind \ind \Stop            &&\to \return \Stop\\
  & \hspace{-11em}\ind \ind \ind \Skip\,\vs       &&\to \return \dlr \Skip\,(\va,\,\vs) \\
  & \hspace{-11em}\ind \ind \ind \Yield\,\vb\,\vs &&\to \return \dlr \Yield\,\vb\,(\va,\,\vs)
\end{alignat*}
We recover let-insertion and case splitting as follows:
\begin{alignat*}{3}
  &\mit{gen_\Pull} : \Lift \vA \to (\Lift \vA \to \Pull\,\vB) \to \Pull\,\vB\\
  &\mit{gen_\Pull}\,\va = \mit{bind_{\Gen}}\,(\gen\,\va)
\end{alignat*}
\begin{alignat*}{3}
  &\mit{case_\Pull} : (\mit{Split}\,\vA,\,\IsSOP\,(\mit{SplitTo}\,\{\vA\})) \RA \Lift \vA \to (\mit{SplitTo}\,\{\vA\} \to \Pull\,\vB) \to \Pull\,\vB\\
  &\mit{case_\Pull}\,\va = \mit{bind_{\Gen}}\,(\mit{split}\,\va)
\end{alignat*}
Let us look at small example. We write $\mit{forEach}$ for $\concatMap$ with
flipped arguments.
\begin{alignat*}{3}
  &\hspace{-4em}\mathrlap{\mit{forEach}\,(\mit{take}\,\qt{100}\,(\mit{countFrom}\,\qt{0})) \dlr \lam \vx.}\\
  &\hspace{-4em}\mathrlap{\gen_\Pull\, \qt{\spl \vx * 2} \dlr \lam \vy.}\\
  &\hspace{-4em}\mathrlap{\mit{case_\Pull}\,\qt{\spl \vx < 50} \dlr \lam \case}\\
  &\hspace{-4em} \ind \True  &&\to \mit{take}\,\vy\,(\mit{countFrom}\,x)\\
  &\hspace{-4em} \ind \False &&\to \single\,\vy
\end{alignat*}
Here, in every $\forEach$ iteration, $\gen_\Pull\,\qt{\spl \vx * 2}$ evaluates
the given expression and saves the result to the machine state. Then,
$\mit{case_\Pull}$ branches on an object-level Boolean expression. If we use
$\foldr$ to generate object code from this definition, we get \emph{four}
mutually defined functions. We get two state shapes from $\single\,\vy$ and one
from $\mit{take}\,\vy\,(\mit{countFrom}\,x)$; we sum these to get three for the
$\mit{case_\Pull}$, then finally we get an extra shape from $\mit{forEach}$ which
introduces an additional $\Maybe$.

@@

\section*{Acknowledgements}
I thank the anonymous reviewers, Sebastian Graf and Jacques Carette, for their
comments and suggestions on draft versions of this paper. This work was supported
by grant 2019.0116 of the Knut and Alice Wallenberg Foundation.

@@

\subsection{Identity Type}
